### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished preliminary draft of parallel chapter. editing is next

%%---------------------------------------------------------------------------%%
\section{Leadership-Class Parallel Scaling Studies\ }
\label{sec:leadership_scaling_studies}
Using the parallel performance metrics presented in the previous
section, we will next performance various scaling studies using an
implementation of the parallel MCSA algorithm to assess the quality of
the algorithm using the neutron diffusion problem presented in
Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
large problems at high levels of concurrency will be solved at a
leadership-class computing facility. In this section, we will outline
the parallel machine on which the calculations were performed and then
present a series of numerical studies that identify both strengths and
weaknesses of the parallel algorithm. Both pure domain decomposition
and MSOD parallel performance will be explored and compared to
performance results for the same problems using a set of production
parallel Krylov solvers. For the scaling results presented here, all
parallel Monte Carlo results were computed using the Monte Carlo
Linear Solvers\footnote{MCLS is an open source library and can be
  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
library generated specifically for this work while the conjugate
gradient and GMRES Krylov solvers used are part of the Belos linear
solver package within the Trilinos framework
\citep{heroux_overview_2005}. In general, the Belos package produces
solutions to the neutron diffusion problem nearly two orders of
magnitude faster than the MCLS library for serial computations using
the conjugate gradient solver. Therefore, when we directly compare
these two codes, the MCLS parallel efficiencies will appear
artificially inflated relative to the Belos efficiencies due to the
fact that the MCLS ratio of work to communication will be much larger
than that for the Belos calculations. For every scaling study
presented, the calculations were performed 3 times with the average
values reported for wall time and used for the efficiency
computations.

\subsection{Titan }
\label{subsec:titan}
All parallel scaling studies presented in this work were performed on
the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
Facility located at Oak Ridge National Laboratory. Titan is a
heterogeneous architecture machine combining both CPU and GPU
components to achieve a theoretical peak performance of more than 20
petaflops. The technical specifications for Titan are given in
Table~\ref{tab:titan_hardware} and specifically for only the CPU
portion of the machine.

@@

For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will isolate
specifically the performance of Brunner and Brantley's algorithm as
applied to the Neumann-Ulam algorithm. For the strong scaling study,
the global problem size was fixed at \sn{1.6}{7} DOFs for all solvers
and the 16 core run used as the base case. MCSA was used again with
one stochastic history used per DOF in the Neumann-Ulam solve to
compute the correction for the acceleration with the remainder of the
problems and solver parameters the same as those for the verification
calculation given in Table~\ref{tab:parallel_verification_parameters}
and Table~\ref{tab:parallel_mcsa_parameters}. As with all of the
scaling studies, 3 calculations for all data points were performed
with the average wall time reported and used for the efficiency
calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the results of the this study
with the given absolute efficiencies computed using
Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
iteration due to the fact that MCSA converged on the solution in 22
iterations while the Krylov solvers converged in 19 iterations.

@@

Beyond around 5,000 cores, the global problem size selected for this
strong scaling study causes the local problem size to shrink enough
that parallel communication costs begin to overtake arithmetic costs
on process. This takeover causes this effective strong scaling wall
that is observed in all methods. Using any more cores to solve this
particular global problem is not an efficient use of those parallel
resources. Furthermore, the fact that we see a very similar trends in
efficiencies for all methods means that they are bound by the same
parallel performance limitations in a strong scaling environment. This
is important as we can expect the same qualitative scaling performance
from MCSA once arithmetic optimization has been performed to improve to
serial performance of the implementation. In addition, these results
are valuable in that the show MCSA, when properly parallelized, has
the ability to be competitive on leadership-class computing platforms
with conventional subspace methods with good strong scaling observed
in the pure domain decomposition case to O(10,000) cores.

@@

As an additional means of comparison between MCSA and the Krylov
methods, we compute the algorithmic efficiencies of the calculations
using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
the Krylov algorithm improves as a function of the number of cores in
the problem due to the fact that the number of iterations required to
converge decreased from 20 in the base case to 19 and then to 18 for
the final set of weak scaling computations. When this reduction in
iterations is used to account for the observed runtimes and computed
absolute efficiencies for weak scaling we end up with the
implementation efficiencies computed using
Eq~(\ref{eq:implementation_efficiency}) given in
Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
same iterative performance as a function of global problem size in the
weak scaling study, the implementation efficiencies compare more
favorably to the Krylov methods. Using the implementation efficiency,
we are effectively measuring the parallel performance of a single
iteration for each method by looking at these implementation
efficiencies. As the global problem size increases, MCSA is less prone
to a drop in parallel efficiency. When optimization has been completed
and runtimes potentially approach those of the Krylov methods, this
comparison may not be as favorable. Overall, weak scaling performance
the parallel MCSA algorithm is excellent for the pure domain
decomposition case with absolute parallel efficiencies of over 80\%
observed at over 32,000 cores.

@@

In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects runtimes and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
\S~\ref{sec:analytic_framework} to design these numerical
experiments. For the neutron diffusion problem used in these scaling
studies, the linear operator had a spectral radius of 0.787 and a
weight cutoff of \sn{1}{-2}. Using Eq~(\ref{eq:analytic_k}) gives an
expected random walk length of 19.22 transitions for any given history
in the problem. Using the mean-chord approximation given by
Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
5.3\% of the histories born in a particular domain to leak out after
the first transport stage. By adding overlap, we aim to reduce this
fraction of communicated histories and therefore boost the parallel
performance of the algorithm. To determine the amount of overlap to
test for the scaling studies, we use a variation of
Eq~(\ref{eq:step_k_length}) to determine how many discrete states a
history will move on average from its birth site. We modify
Eq~(\ref{eq:step_k_length}) to only account for discrete movements:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
discrete states a history will move from its birth site. For our
diffusion problem used in these scaling studies, we compute a value of
2.63 states per history for any of the dimensions of the system. This
means that we can expect to end up on average 2.63 states in both the
$i$ and $j$ dimensions away from the starting state. Therefore, we
will use the overlap generation algorithm outlined in
\S~\ref{subsec:msod_generation} such that we parameterically grow the
local domain using 1, 2, 3, and 5 levels to measure how scalability is
affected when this root-mean-squared number of states is
considered. For example, using 3 levels of overlap should keep over
50\% of the histories that would be communicated in the pure domain
decomposition case on-process as this is greater than the
root-mean-squared number of states we would expect a history to
travel.

@@

@@

@@

@@

We next look to apply replication to the problem in the form of
multiple sets. Before performing parallel scaling studies using
multiple sets, several aspects of the replicated parallel problem must
be considered in order to effectively craft the studies and interpret
their results. First is the question of how many stochastic histories
to use to generate the correction vector in the MCSA iteration. For
this work we consider two choices; the split case and the fully
replicated case. As an example these cases, consider a single set
problem where 20,000 histories are used in the Monte Carlo solve. If
we choose the split case, then solving the same problem with 2 sets
would give 10,000 histories in each set and 4 sets would give 5,000
histories in each set. Splitting the histories in this way provides
the same global number of histories used to compute the correction in
the single set problem. We expect to reduce the time to solution by
doing this as each individual set solve will occur faster in an almost
linear fashion with the reduced number of histories. For the
replicated history case, each additional set in the problem will use
just as many histories as the single set case. Using 2 sets with fully
replicated histories for our example problem then has each set
computing 20,000 histories with 40,000 global histories in the
computation. For the 4 set case, the number of global histories is
increased to 80,000. We also expect this type of replication to reduce
the time to solution as the extra global histories should reduce the
stochastic uncertainty in the correction vector and reduce the overall
number of MCSA iterations required to converge.

@@

Also of interest here is the effect of splitting histories among
sets versus replicating them on the iterative performance of the
method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
required to converge and computed algorithmic efficiencies for each of
the cases presented in the strong scaling data. As expected, splitting
histories maintains a fixed global number of histories and therefore
maintained the number of iterations required to converge relative to
the single set base case. When histories were replicated, adding sets
decreased the number of iterations required to converge in a nonlinear
fashion. Using these algorithmic efficiencies, the implementation
efficiencies for the multiple set cases were computed and are given in
Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
that the replicated history cases are even less efficient than in the
absolute case. Not only do we have the additional parallel overhead of
doing the block level reduction, but we also have additional histories
that increase the time to solution for a single iteration.

@@

Next, we continue to a weak scaling study of the multiple sets
problem. In the same way as the strong scaling study, we will modify
the pure domain decomposition study to account for the additional
replication. For each calculation in the single set study, that
problem will be replicated either 2 or 4 times with both split and
replicated histories to assess the effects on scalability and
runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
the multiple sets weak scaling study. Of primary importance here is
the fact that adding sets to the problem with both split and
replicated histories actually decreases the time to solution. This was
not observed in the strong scaling case until much larger core counts
due as the multiple set computations hit the strong scaling wall after
the single set computation.

@@

Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
scaling efficiencies using the multiple set metric given by
Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
solution over the base cases for nearly all data points, the
observation here is that the reduction in time to solution is still
not large enough to efficiently use all resources. For example, for
the 2 set case to be a perfect use of resources with either split or
replicated histories, the runtime of that calculation must be half of
that for the single set case. For the 4 set case, this run time should
be a quarter of the single set case for perfect scaling. This was not
the observation for the runtimes in
Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
is still the most efficient use of computational resources in this
scaling study.

@@

As in the strong scaling case, splitting and replicating histories
across sets had effectively the same results for the weak scaling
study with the iterations to converge and computed algorithmic
efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
implementation efficiencies are computed, we see in
Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
sets is still a more efficient manner of introducing multiple sets
rather than replicating histories in an attempt to reduce the number
of iterations required to converge.  Here we also note that this is
not a true weak scaling study as indicated by
Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
weak and strong scaling. Regardless of this, although multiple sets
are not the most efficient use of resources in this analysis, they do
enhance the time to solution through replication and do so
significantly at core counts of O(1,000) and lower. In both the weak
and strong scaling cases, due to the additional computational and
communication overhead of adding sets to the system, one should never
expect to seriously boost parallel efficiencies for MCSA using this
technique accept for certain instances past the strong scaling wall in
a purely strong scaling environment. In a weak scaling environment,
although no efficiency increase will be noted, one can expect an
enhancement of time to convergence by replicating.

@@

Based on the analysis of the pure domain decomposed MCSA algorithm and
subsequently the MSOD implementation, one may ask then ask if there is
any need to communicate histories from domain-to-domain at
all. Overlap was not as successful because so few histories needed to
be communicated in the first place that keeping them on-process
resulted in more work than the communications savings that
occurred. If that is true, then perhaps without any overlap or
communication, just doing local Monte Carlo may be satisfactory for
accelerating the MCSA iteration. In this case, histories near the
boundaries of the domain may be truncated before reaching the weight
cutoff if the reach the domain boundary first. This truncation has the
potential to increase the stochastic uncertainty of the Neumann-Ulam
result and therefore increase the number of MCSA iterations required
to converge. In addition, the scalability of the method will then be
purely limited by the parallel matrix and vector operations required
to implement the remainder of the algorithm.

Considering using MCSA in this way such that the Monte Carlo solve
occurs only in the subdomains casts the method as more of a true
domain decomposition as outlined by Chapter 14 of Saad's text on
iterative methods \citep{saad_iterative_2003}. Compared to traditional
domain decomposition methods we see many similarities including the
potential to overlap the subdomains to improve the results and the
approximate subdomain solves effectively through a relaxation
scheme. By building the boundary data as outlined in
\S~\ref{subsec:domain_generation}, we are effectively gathering one
level of the adjacent domains, albeit in this case to satisfy the data
requirements of the Monte Carlo estimators. Furthermore, the fact that
we are using residual Monte Carlo to compute the correction vector
casts MCSA as a stochastic realization of an additive Schwarz method
with a fixed point iterations used as a relaxation step\footnote{See
  section 14.3.3 in \citep{saad_iterative_2003} for a full definition
  of the additive Schwarz method.}. Both the fixed point iteration and
the computation of the residual propagate the solution from individual
subdomains to the others through the matrix-vector multiply operation
after all subdomain Monte Carlo solves have been completed. Perhaps an
even more important consequence of viewing MCSA as a stochastic
additive Schwarz procedure is that it has the potential to more
readily fit as an accelerator into other linear solver techniques
including GMRES or potentially as a preconditioner.

Using the Neumann-Ulam solver to compute the MCSA correction only in
the subdomains, the strong and weak scaling studies were repeated to
assess the performance gains made by eliminating communication in the
Monte Carlo solver. For each scaling study, the 16-core base case with
Monte Carlo communication was used as the reference point for
efficiency computations in order to measure how eliminating
communication boosts
efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
results of the strong scaling study and
Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
weak scaling study in terms of absolute efficiencies. In both cases,
the benefits to parallel performance of using MCSA as a subdomain
solver are obvious. In the strong scaling case we make the important
observation that we hit the strong scaling wall at the same time with
and without the additional Monte Carlo communication. Although the
Monte Carlo communication did contribute to the efficiency decrease,
it is clear that the remaining matrix and vector operations are the
primary impediment to scalability. Furthermore, we see an even greater
rise above unity in efficiencies because there is less communication
overhead hiding the better cache utilization as the local problem
shrinks. At 16,384 cores, performing the subdomain solves boost the
absolute efficiency from 38\% to 67\% relative to the reference case.

>>>>

%%---------------------------------------------------------------------------%%
\section{Leadership-Class Parallel Scaling Studies\ }
\label{sec:leadership_scaling_studies}
Using the parallel performance metrics presented in the previous
section, we will next measure performance in various scaling studies
using an implementation of the parallel MCSA algorithm to assess its 
quality using the neutron diffusion problem presented in
Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
large problems at high levels of concurrency will be solved at a
leadership-class computing facility. In this section, we will outline
the parallel machine on which the calculations were performed and then
present a series of numerical studies that identify both strengths and
weaknesses of the parallel algorithm. Both pure domain decomposition
and MSOD parallel performance will be explored and compared to
performance results for the same problems using a set of production
parallel Krylov solvers.

For the scaling results presented here, all parallel Monte Carlo
results were computed using the Monte Carlo Linear
Solvers\footnote{MCLS is an open source library and can be downloaded
  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
specifically for this work while the conjugate gradient and GMRES
Krylov solvers used are part of the Belos linear solver package within
the Trilinos framework \citep{heroux_overview_2005}. In general, the
Belos package produces solutions to the neutron diffusion problem
nearly two orders of magnitude faster than the MCLS library for serial
computations using the conjugate gradient solver. Therefore, when we
directly compare these two codes, the MCLS parallel efficiencies will
appear artificially inflated relative to the Belos efficiencies due to
the fact that the MCLS ratio of work to communication will be much
larger than that for the Belos calculations. For every scaling study
presented, the calculations were performed 3 times with the average
values reported for wall time and used for the efficiency
computations. In addition, it should be noted that in practice,
symmetric positive-definite problems like the neutron diffusion
problem solved here would never be solved using GMRES. Instead, a
method such as conjugate gradient would typically be used due to the
enhanced performance. We use GMRES here to simply as a means of
comparison to MCSA performance.

\subsection{Titan }
\label{subsec:titan}
All parallel scaling studies presented in this work were performed on
the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
Facility located at Oak Ridge National Laboratory. Titan is a
heterogeneous architecture machine combining both CPU and GPU
components to achieve a theoretical peak performance of more than 20
petaflops. The technical specifications for Titan for only the CPU
portion of the machine are given in Table~\ref{tab:titan_hardware}.

@@

For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will isolate
specifically the performance of Brunner and Brantley's algorithm as
applied to the Neumann-Ulam algorithm. For the strong scaling study,
the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
all solvers and the 16 core run used as the base case. MCSA was used
again with one stochastic history used per DOF in the Neumann-Ulam
solve to compute the correction for the acceleration with the
remainder of the problems and solver parameters the same as those for
the verification calculation given in
Table~\ref{tab:parallel_verification_parameters} and
Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
studies, 3 calculations for all data points were performed with the
average wall time reported and used for the efficiency
calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the results of the this study
with the given absolute efficiencies computed using
Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
iteration due to the fact that MCSA converged on the solution in 22
iterations while the Krylov solvers converged in 19 iterations.

@@

Beyond around 5,000 cores, the global problem size selected for this
strong scaling study causes the local problem size to shrink enough
that parallel communication costs begin to overtake arithmetic costs
on process. This takeover causes this effective strong scaling wall
that is observed in all methods. Using any more cores to solve this
particular global problem is not an efficient use of those parallel
resources. Furthermore, the fact that we see a very similar trends in
efficiencies for all methods means that they are bound by the same
parallel performance limitations in a strong scaling environment. This
is important as we can expect the same qualitative scaling performance
from MCSA if arithmetic optimization has been performed to improve the
serial performance of the implementation. In addition, these results
are valuable in that they show MCSA, when properly parallelized, has
the ability to be competitive on leadership-class computing platforms
with conventional subspace methods with good strong scaling observed
in the pure domain decomposition case to O(10,000) cores.

@@

As an additional means of comparison between MCSA and the Krylov
methods, we compute the algorithmic efficiencies of the calculations
using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
the Krylov algorithm improves as a function of the number of cores in
the problem due to the fact that the number of iterations required to
converge decreased from 20 in the base case to 19 and then to 18 for
the final set of weak scaling computations. When this reduction in
iterations is used to account for the observed runtimes and computed
absolute efficiencies for weak scaling we end up with the
implementation efficiencies computed using
Eq~(\ref{eq:implementation_efficiency}) given in
Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
same iterative performance as a function of global problem size in the
weak scaling study, the implementation efficiencies compare more
favorably to the Krylov methods. Using the implementation efficiency,
we are effectively measuring the parallel performance of a single
iteration for each method by looking at these implementation
efficiencies. As the global problem size increases, MCSA is less prone
to a drop in parallel efficiency. If optimization has been performed
and runtimes potentially approach those of the Krylov methods, this
comparison may not be as favorable. Overall, weak scaling performance
of the parallel MCSA algorithm is excellent for the pure domain
decomposition case with absolute parallel efficiencies of over 80\%
observed at over 32,000 cores.

@@

In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects runtimes and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
\S~\ref{sec:analytic_framework} to design these numerical
experiments. For the neutron diffusion problem used in these scaling
studies, the linear operator had a spectral radius of 0.787 and a
weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
length of 19.22 transitions for any given history in the
problem. Using the mean-chord approximation given by
Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
5.3\% of the histories born in a particular domain to leak out after
the first transport stage. By adding overlap, we aim to reduce this
fraction of communicated histories and therefore boost the parallel
performance of the algorithm.

To determine the amount of overlap to test for the scaling studies, we
use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
discrete states a history will move on average from its birth site. We
modify Eq~(\ref{eq:step_k_length}) to only account for discrete
movements:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
discrete states a history will move from its birth site. For our
diffusion problem used in these scaling studies, we compute a value of
2.63 states per history for any of the dimensions of the system. This
means that we can expect to end up on average 2.63 states in both the
$i$ and $j$ dimensions away from the starting state. Therefore, we
will use the overlap generation algorithm outlined in
\S~\ref{subsec:msod_generation} such that we parameterically grow the
local domain using 1, 2, 3, and 5 levels to measure how scalability is
affected when this root-mean-squared number of states is
considered. For example, using 3 levels of overlap should keep over
50\% of the histories that would be communicated in the pure domain
decomposition case on-process as this is greater than the
root-mean-squared number of states we would expect a history to
travel.

@@

The absolute efficiency results from the strong scaling exercise are
given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
boosting of efficiencies at 2,048 cores, little evidence of strong
scaling improvement is evident for all levels of overlap tested. To
further explore any potential benefits of adding overlap into the
system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
difference in the efficiencies between all overlap cases and the
reference case without overlap at each core count. At smaller numbers
of cores, there is no observed benefit to using overlap to improve
strong scaling. The improvements noted at 2,048 cores are also present
here and account for the best efficiency improvements for all core
counts tested. at 16,384 cores, using 2 levels of overlap did give a
6.4\% boost in efficiency while at 1,024 cores this same amount of
overlap actually reduced the efficiency. Beyond 2 levels of overlap,
efficiency levels begin to fall with 5 levels of overlap performing
the worst for all cases. From this, we don't expect adding any more
overlap will improve the scalability of the algorithm and the best
strong scaling performance was observed when the number of overlap
levels is approximately equal to the root-mean-squared number of
states we expect a history to move from its starting point as computed
by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
the drop in efficiencies for levels of overlap beyond this value until
we have analyzed the weak scaling data.

@@

For weak scaling, the same study as in the pure domain decomposition
case was performed again with 1, 2, 3, and 5 levels of overlap. The
absolute parallel efficiencies from these calculations are given in
Figure~\ref{fig:titan_weak_overlap} and the differences in
efficiencies from the calculations without overlap. The results here
are much the same in that there is not a dramatic improvement in
scalability relative to the case without overlap. Opposite the strong
scaling calculation, overlap enhances weak scaling efficiencies at
lower core counts and degrades the efficiencies at higher core
counts. In addition, 1 level of overlap provided the best enhancement
of scalability when overlap was beneficial.

@@

For both the strong and weak scaling studies, the parallel performance
of the algorithm was either marginally improved or not improved at
all. We would always expect an improvement in performance for using
overlap to prevent the transport of histories from domain to domain
during execution of the Neumann-Ulam algorithm. The reduction in
tranport means a reduction in parallel communication and therefore
should never degrade the performance of the algorithm. It is true that
doing so in fact enhances Monte Carlo performance. However, the
Neumann-Ulam algorithm is not representative of the entire MCSA
algorithm. When overlap is used, an overlapping tally vector is
constructed in parallel such that multiple cores may own multiple
states in the system. When this occurs, the tally data associated with
these shared states must be communicated after the Neumann-Ulam solve
as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
operation then builds the full MCSA correction vector in the
decomposition of the solution vector.

Therefore, by building overlap we are in fact reducing parallel
communication during the Neumann-Ulam solve but we are simply
deferring that communication until after Monte Carlo transport when
the overlapping tally vector must be reduced. When overlap becomes
large enough (i.e. larger than the root-mean-squared number of
discrete states a history will travel), then the savings in
communication during the Neumann-Ulam solve start to become smaller
than the communication created in the overlapping tally vector
reduction. We see this in both the strong and weak scaling cases at
larger levels of overlap. For the strong scaling case, we see better
performance improvement at higher core counts because the total number
of overlapping states on each core is smaller from the local problem
size shrinking. For weak scaling, we see better performance
improvement at lower numbers of cores because although the local
overlap size is fixed, the global amount of overlap is growing with
core count and therefore reducing the performance of the tally
reduction operation.

Based on this data, a few levels of overlap within the expected
distance of travel for a history may offer some marginal improvement
in scalability, however, it is not as effective as one might think
given its improvement to reactor physics calculations as observed in
the literature. This is largely due to the density of states in the
system that creates the described performance bottlenecks in the
overlapping tally vector reduction operation. In a Monte Carlo
particle simulation, these overlapping tally reduction operations are
not expected to be peformed at the speed they are required in
MCSA. For example, in a criticality calculation with overlapping
domains, we would expect to perform as many parallel tally vector
reduction in a particle transport simulation as there are particle
stages required for convergence of the eigenvalue. However, these
particle stages will likely be much more costly in terms of wall time
relative to a Neumann-Ulam solve in an MCSA iteration due to the more
extensive particle physics and logic structure that must be
evaluated. Therefore, we can expect this overlapping tally reduction
to be less signficant to scalability in those operations. We do not
have as complicated a logic structure in a history random walk for the
Neumann-Ulam method and therefore overlap can inhibit scalability if
too large.

@@

We next look to apply replication to the problem in the form of
multiple sets. As the previous analysis showed overlap to be of little
assistance in improving scalability, we only apply multiple sets to
the problem with pure domain decomposition within the sets for these
scaling studies. Before performing parallel scaling studies using
multiple sets, several aspects of the replicated parallel problem must
be considered in order to effectively craft the studies and interpret
their results. First is the question of how many stochastic histories
to use to generate the correction vector in the MCSA iteration. For
this work we consider two choices; the split case and the fully
replicated case. As an example of these cases, consider a single set
problem where 20,000 histories are used in the Monte Carlo solve. If
we choose the split case, then solving the same problem with 2 sets
would give 10,000 histories in each set and 4 sets would give 5,000
histories in each set. Splitting the histories in this way provides
the same global number of histories used to compute the correction in
the single set problem. We expect to reduce the time to solution by
doing this as each individual set solve will occur faster in an almost
linear fashion with the reduced number of histories. For the
replicated history case, each additional set in the problem will use
just as many histories as the single set case. Using 2 sets with fully
replicated histories for our example problem then has each set
computing 20,000 histories with 40,000 global histories in the
computation. For the 4 set case, the number of global histories is
increased to 80,000. We also expect this type of replication to reduce
the time to solution as the extra global histories should reduce the
stochastic uncertainty in the correction vector and reduce the overall
number of MCSA iterations required to converge.

@@

Also of interest here is the effect of splitting histories among sets
versus replicating them on the iterative performance of the
method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
required to converge and computed algorithmic efficiencies for each of
the cases presented in the strong scaling data. As expected, splitting
histories maintains a fixed global number of histories and therefore
maintained the number of iterations required to converge relative to
the single set base case. When histories were replicated, adding sets
decreased the number of iterations required to converge in a nonlinear
fashion. Therefore, increasing the number of histories per MCSA
iteration improves algorithmic efficiency. Using these algorithmic
efficiencies, the implementation efficiencies for the multiple set
cases were computed and are given in
Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
that the replicated history cases are even less efficient than in the
absolute case with respect to the performance of a single
iteration. Not only do we have the additional parallel overhead of
doing the block level reduction, but we also have additional histories
that increase the time to solution for a single iteration.

@@

Next, we continue to a weak scaling study of the multiple sets
problem. In the same way as the strong scaling study, we will modify
the pure domain decomposition study to account for the additional
replication. For each calculation in the single set study, that
problem will be replicated either 2 or 4 times with both split and
replicated histories to assess the effects on scalability and
runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
the multiple sets weak scaling study. Of primary importance here is
the fact that adding sets to the problem with both split and
replicated histories actually decreases the time to solution. This was
not observed in the strong scaling case until much larger core counts
as the multiple set computations hit the strong scaling wall after the
single set computation.

@@

Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
scaling efficiency computations using the multiple set metric given by
Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
solution over the base cases for nearly all data points, the
observation here is that the reduction in time to solution is still
not large enough to efficiently use all resources. For example, for
the 2 set case to be a perfect use of resources with either split or
replicated histories, the runtime of that calculation must be half of
that for the single set case. For the 4 set case, this run time should
be a quarter of the single set case for perfect scaling. This was not
the observation for the runtimes in
Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
is still the most efficient use of computational resources in this
scaling study.

@@

As in the strong scaling case, splitting and replicating histories
across sets had effectively the same results for the weak scaling
study with the iterations to converge and computed algorithmic
efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
implementation efficiencies are computed, we see in
Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
sets is still a more efficient manner of introducing multiple sets
rather than replicating histories in an attempt to reduce the number
of iterations required to converge.  Here we also note that this is
not a true weak scaling study as indicated by
Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
weak and strong scaling. Regardless of this, although multiple sets
are not the most efficient use of resources in this analysis, they do
enhance the time to solution through replication and do so
significantly at core counts of O(1,000) and lower. In both the weak
and strong scaling cases, due to the additional computational and
communication overhead of adding sets to the system, one should never
expect to seriously boost parallel efficiencies for MCSA using this
technique except for certain instances past the strong scaling wall in
a purely strong scaling environment. In a weak scaling environment,
although no efficiency increase will be noted, one can expect an
enhancement of time to convergence by replicating.

@@

Based on the analysis of the pure domain decomposed MCSA algorithm and
subsequently the multiple sets and overlapping domain implementation,
one may ask then ask if there is any need to communicate histories
from domain-to-domain at all. Overlap was not as successful because
because the additional communication required to reduce the
overlapping tally vector eventually becomes larger than the
communication savings during Monte Carlo transport. In addition, for
local domains that are large enough to get good performance
(e.g. those used in the weak scaling studies), then our analytic
relationships for domain leakage, and in particular
Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
of the histories in this diffusion problem will leave the domain in
the first place.  If that is true, then perhaps without any overlap or
communication, just doing local Monte Carlo transport such that
histories are terminated upon reaching the domain boundary may be
satisfactory for accelerating the MCSA iteration. In this case,
histories near the boundaries of the domain may be truncated before
reaching the weight cutoff if they reach the domain boundary
first. This truncation has the potential to increase the stochastic
uncertainty of the Neumann-Ulam result and therefore increase the
number of MCSA iterations required to converge. In addition, the
scalability of the method will then be purely limited by the parallel
matrix and vector operations required to implement the remainder of
the algorithm on the subset of processors containing the linear
problem.

Considering using MCSA in this way such that the Monte Carlo solve
occurs only in the subdomains casts the method as more of a true
domain decomposition as outlined by Chapter 14 of Saad's text on
iterative methods \citep{saad_iterative_2003}. Compared to traditional
domain decomposition methods we see many similarities including the
potential to overlap the subdomains to improve the results. By
building the boundary data as outlined in
\S~\ref{subsec:domain_generation}, we are effectively gathering one
level of the adjacent domains, albeit in this case to satisfy the data
requirements of the Monte Carlo estimators. Furthermore, the fact that
we are using residual Monte Carlo to compute the correction vector
casts MCSA as a stochastic realization of an additive Schwarz method
with a fixed point iteration used as a relaxation step\footnote{See
  section 14.3.3 in \citep{saad_iterative_2003} for a full definition
  of the additive Schwarz method.}. Both the fixed point iteration and
the computation of the residual propagate the solution from individual
subdomains to the others through the matrix-vector multiply operation
after all subdomain Monte Carlo solves have been completed. Perhaps an
even more important consequence of viewing MCSA as a stochastic
additive Schwarz procedure is that it has the potential to more
readily fit as an accelerator into other linear solver techniques
including GMRES or potentially as a preconditioner.

Using the Neumann-Ulam solver to compute the MCSA correction only in
the subdomains, the strong and weak scaling studies were repeated to
assess the performance gains made by eliminating communication in the
Monte Carlo solver. For each scaling study, the 16-core base case with
Monte Carlo communication was used as the reference point for
efficiency computations in order to measure how eliminating
communication boosts
efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
results of the strong scaling study and
Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
weak scaling study in terms of absolute efficiencies. In both cases,
the benefits to parallel performance of using MCSA as a subdomain
solver are obvious. In the strong scaling case we make the important
observation that we hit the strong scaling wall at the same time with
and without the additional Monte Carlo communication. Although the
Monte Carlo communication did contribute to the efficiency decrease,
it is clear that the remaining matrix and vector operations are the
primary impediment to scalability. Furthermore, we see an even greater
rise above unity in efficiencies because there is less communication
overhead hiding the better cache utilization as the local problem size
shrinks. At 16,384 cores, performing the subdomain solves boost the
absolute efficiency from 38\% to 67\% relative to the reference case.
